A Study on Common Bug Fix Patterns in Rust

### 2022-09-23 sections/methodology.tex
added essence figure, revised methodology, added introduction for results, revised tables for results, we still need to add some numbers

% Let's start this by talking right away about changes rather than just ASTs
The general idea for automatically finding the pervasive code patterns within a list of programs is to parse the programs and encode the important parts of the derived AST in fixed-size datapoints. We need fixed-size datapoints, as it is a requirement for both centroid-based and density-based well-known clustering algorithms. Work in this vein makes the assumption that each cluster corresponds to one change pattern.

@@

% TODO Define change
A change pattern categorizes a class of changes. We thus need to define what a change is. For our purposes, a change is ....
To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs, encoded in two Python dictionaries. Tree diff algorithms can compute the differences between two arbitrary trees. If the trees are programs' abstract syntax trees, we call their diff an ASTDiff. We use dictdiff to obtain ASTDiffs.

@@

% TODO: BCL elements goes in an appendix.
As mentioned above, we use Syn for parsing Rust. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related lexemes (BCL) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \verb+BCL+ elements can be found within our artifact. 
% I think we don't need this sentence here. Let's see what we do later.
%We define dimension items as \verb+DI = NT+ $\cup$ \verb+BCL+, as this set also equals the columns of our tables.

@@

An ASTDiff, as computed by dictdiff, contains a type:
\begin{itemize}
\item `add': A new structure has been added to the tree; 
\item `remove': A structure has been dropped from the tree; 
\item `change': The content of a sub structure of the first tree has changed.
\end{itemize}
% the description of the context in 2.1.2 is probably enough, you don't need it here, just remove it. I think that you just put the above itemize into 2.1.2 and erase the next paragraph, unless you can spot something you need to add to 2.1.2.
It also encodes the context of the modification. This context happens to be the root elements in our file, e.g. function blocks, struct blocks, impl blocks, etc. The modified content within a context is still in the form of tree. Our objective right now is to find primary elements in our diff that describe the semantics of occurred change.
% TODO rework the above paragraph

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{figs/extraction.png}
\caption{\label{fig:extraction}Path Extraction}
\end{figure}

% remove redundancy between the description of the ASTDiff in 2.1.1 and what we have here. Probably it can go all into 2.1.2, i.e. you can just move the last para of 2.1.1 to here and incorporate it.
The output of dictdiff is a list of diffs, where each diff comprises three parts. The first part specifies whether the modification is of type `add', `remove', or `change'. The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change (terminals and nonterminals). Our tool looks for changes within each root-level scope. That is, our assumption is that a change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool would detect both continuous and non-continuous line changes within one scope.
% TODO provide an example and then transform it into fig:extraction.

@@

% I think I would move the discussion of why fixed size from the beginning of the section to here; we don't need it at the beginning of the section.
% I think it would be better to show an example of the two if statements either way
\paragraph{Encoding the data.} A key question is how to transform these paths to a fixed size datapoint. First, we need to define a fixed set of columns. The number of columns would depend on the amount of information we want to encode in the datapoints. The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an encoding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to encode all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCL}$ as our dimensions, which yields an infeasible $O(n^n)$ dimensions; this could potentially be reduced somewhat, but is still impractical. A reasonable work around would be to reduce $n$ from the full set of non terminals to a smaller set of categories (e.g. a category for larger entites like class or function definitions, and a category for smaller entities like statements and expressions), and then to record all possible combinations of these categories (similar to JS authors~\cite{foo}). That approach reduces the number of dimensions and provides more reasonable and syntactically-correct combinations. However, it would still yield a sparse dataset. 
% todo you need to dig up the JS paper ref

% ok, add a figure and then we'll look at the text again and follow along with the picture
In our methodology, we propose a novel representation. Similar to the first approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCL}$ we collected from Syn. However, to simulate order-sensitiveness, we carried out two auxiliary steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

@@

We made two observations about changes. First, we saw that an item (non terminal or BC-related lexeme) that occurs closer to AST leaves tends to be more semantically important. Relatedly, items closer to the root tend to be repeated in all the paths within one ASTDiff (e.g. ItemFn, Local, ExprCall in Figure~\ref{fig:extraction}). We thus want to prioritize items that tend to occur closer to leaves and de-prioritize items that tend to occur closer to roots.

@@

@@

\begin{table}[h]
\centering
\caption{\label{table:repos} Target Repositories}
\includegraphics[width=1\textwidth]{figs/repos.png}
\end{table}

% should have a reference to how you are computing churn
Now that we have set up our code analysis pipeline, we can start mining the Rust repositories. We target the top 18 most-starred Rust projects on GitHub, at the time of data collection. Table~\ref{table:repos} summarizes our benchmarks. Because we are interested in changes, we computed the average code churn of the files within the projects. denoland/deno and swc-project/swc are the projects with largest number of LOCs and highest average code churn, respectively. Unsurprisingly, we captured a lot of instances from these two projects in our final clusters.

We used the Pydriller library to mine software repositories. Pydriller provides APIs to extract commits from a Git repository and to search through different revisions of files. Algorithm~\ref{alg} shows how we used Pydriller to run the target repositories through our code analysis pipeline and populate two databases: one for borrow-checker related patterns (which involve items in \texttt{BCL}) and one for general patterns (which don't). For each repository, we search all commits that include bug-fixing related keywords within their commit messages. Here, we chose bug-fixing related keywords `bug', `error', `fault', `defect', and `fix'. 
% cite earlier work that does the search-for-keywords-to-identify-bugs; my earlier EMSE paper has that citation for instance.

@@

% TODO fix this caption
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{figs/clusters.png}
\caption{\label{fig:clustering} Clustering Experiments}
\end{figure}

@@

In this work, we were only interested in the first two groups and ignored the clusters for refactoring changes. Also, we prefer reporting clusters that manifest concrete patterns. An abstract pattern as opposed to a concrete pattern can account for many changes in the programs. For instance, the pattern `Adding a new statement to a function's body' is deemed more abstract than `Changing a clone of a variable to a borrowing of it in a function argument'. % show example in terms of code; do you have an actual definition for abstract vs concrete?

>>>>

A code change is a modification in a program that changes the program's abstract syntax tree. A change pattern is a series of code changes that serves a specific purpose. This purpose can be improving code readability or maintainability (refactoring patterns), or rectifying program behaviour with respect to a specific functional or non-functional requirement (bug-fixing patterns). In this work, we are interested in latter group. Thus, from now on, by referring to code patterns we mean bug-fixing patterns.

The general idea for automatically finding the pervasive code patterns within a list of programs is to parse the programs and encode the important parts of the derived AST in fixed-size datapoints. Work in this vein makes the assumption that each cluster corresponds to one change pattern.

@@

To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs, encoded in two Python dictionaries. Tree diff algorithms can compute the differences between two arbitrary trees. If the trees are programs' abstract syntax trees, we call their diff an ASTDiff. We use dictdiff to obtain ASTDiffs.

@@

As mentioned above, we use Syn for parsing Rust. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related lexemes (BCL) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \verb+BCL+ elements can be found within Appendix~\ref{sec:appendix}. 
% I think we don't need this sentence here. Let's see what we do later.
%We define dimension items as \verb+DI = NT+ $\cup$ \verb+BCL+, as this set also equals the columns of our tables.

@@

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{figs/extraction.png}
\caption{\label{fig:extraction}Extracting paths from an ASTDiff. AST corresponds to the code change in the yellow box}
\end{figure}

The output of dictdiff is a list of diffs, where each diff comprises three parts. The first part specifies the modification type:

\begin{itemize}
    \item `add': A new structure has been added to the tree; 
    \item `remove': A structure has been dropped from the tree; 
    \item `change': The content of a sub structure of the first tree has changed.
\end{itemize}

The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change (terminals and nonterminals). Our tool looks for changes within each root-level scope. That is, our assumption is that a change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool would detect both continuous and non-continuous line changes within one scope.

@@

\paragraph{Encoding the data.} A key question is how to transform these paths to a fixed size datapoint. First, we need to define a fixed set of columns. We need fixed-size datapoints, as it is a requirement for both centroid-based and density-based well-known clustering algorithms. The number of columns would depend on the amount of information we want to encode in the datapoints. The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an encoding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to encode all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCL}$ as our dimensions, which yields an infeasible $O(n^n)$ dimensions; this could potentially be reduced somewhat, but is still impractical. However, in this case we can put a distinction between the aformentioned if statements, as we have a different dimension for nested if statements (\verb+if_if_statement_...+). A reasonable work around would be to reduce $n$ from the full set of non terminals to a smaller set of categories (e.g. a category for larger entites like class or function definitions, and a category for smaller entities like statements and expressions), and then to record all possible combinations of these categories (similar to JS authors~\cite{foo}). That approach reduces the number of dimensions and provides more reasonable and syntactically-correct combinations. However, it would still yield a sparse dataset. 
% todo you need to dig up the JS paper ref / Not sure, how to cite, I usually create a bib file, but i'm not sure if i'm allowed to do that for this template.

In our methodology, we propose a novel representation. Similar to the first approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCL}$ we collected from Syn. However, to simulate order-sensitiveness, we carried out two auxiliary steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{figs/essence.png}
    \caption{\label{fig:essence}Multiplying the number of occurrences of items observed in the tree by their respective weight, resulting in the essence of change (Visualized with a circle pack figure)}
\end{figure}

@@

We made two observations about changes. First, we saw that an item (non terminal or BC-related lexemes) that occurs closer to AST leaves tends to be more semantically important. Relatedly, items closer to the root tend to be repeated in all the paths within one ASTDiff (e.g. ItemFn, Local, ExprCall in Figure~\ref{fig:extraction}). We thus want to prioritize items that tend to occur closer to leaves and de-prioritize items that tend to occur closer to roots.

@@

Multiplying the number of occurrences of all observed items in a change by their respective weight results in a vector of numbers. We call this vector, the essence of change. In this vector, the respective value for each item shows its importance in the change. Figure~\ref{fig:essence} shows the process of computing the essence of the code change that we saw in Figure~\ref{fig:extraction}. We have visualized this essence vector using a circle pack figure. The radius of each circle corresponds to the importance of the item inside it.

@@

\begin{table}
\caption{\label{table:repos} Target repositories for mining}
\begin{tabular}{|c|c|c|c|c|}
    \hline
    Project Name & Stars(k) & LOC & \# Commits & Avg. Code Churn \\ \hline
    denoland/deno & 84.7 & 562468 & 7730 & 49.22 \\ \hline
    tauri-apps/tauri & 48.5 & 130726 & 2929 & 31.67 \\ \hline
    alacritty/alacritty & 41.2 & 45803 & 2017 & 27.53 \\ \hline
    sharkdp/bat & 36.4 & 93258 & 2392 & 59.13 \\ \hline
    starship/starship & 28 & 176630 & 2290 & 25.1 \\ \hline
    rust-lang/rustlings & 28.7 & 9852 & 1391 & 8.92 \\ \hline
    meilisearch/meilisearch & 28.6 & 66872 & 3849 & 279.7 \\ \hline
    sharkdp/fd & 24.2 & 10383 & 1025 & 12.2 \\ \hline
    swc-project/swc & 23.5 & 13512766 & 5205 & 44.34 \\ \hline
    yewstack/yew & 23.2 & 112972 & 2239 & 21.79 \\ \hline
    xi-editor/xi-editor & 19.5 & 58783 & 2104 & 28.67 \\ \hline
    AppFlowy-IO/AppFlowy & 25.9 & 182830 & 2600 & 29.49 \\ \hline
    firecracker-microvm/firecracker & 18.9 & 156759 & 3536 & 58.26 \\ \hline
    nushell/nushell & 20.3 & 181710 & 5922 & 22.38 \\ \hline
    ogham/exa & 18.9 & 17114 & 1558 & 11.99 \\ \hline
    SergioBenitez/Rocket & 18.2 & 90055 & 2127 & 19.75 \\ \hline
    rustdesk/rustdesk & 27.7 & 85322 & 1664 & 72.81 \\ \hline
    tokio-rs/tokio & 17.2 & 126089 & 3006 & 28.87 \\ \hline
\end{tabular}
\end{table}

Now that we have set up our code analysis pipeline, we can start mining the Rust repositories. We target the top 18 most-starred Rust projects on GitHub, at the time of data collection. Table~\ref{table:repos} summarizes our benchmarks. Because we are interested in changes, we computed the average code churn of the files within the projects (The code churn is computed using process metrics provided by Pydriller). denoland/deno and swc-project/swc are the projects with largest number of LOCs and highest average code churn, respectively. Unsurprisingly, we captured a lot of instances from these two projects in our final clusters.

We used the Pydriller library to mine software repositories. Pydriller provides APIs to extract commits from a Git repository and to search through different revisions of files. Algorithm~\ref{alg} shows how we used Pydriller to run the target repositories through our code analysis pipeline and populate two databases: one for borrow-checker related patterns (which involve items in \texttt{BCL}) and one for general patterns (which don't). For each repository, we search all commits that include bug-fixing related keywords within their commit messages. Here, we chose bug-fixing related keywords `bug', `error', `fault', `defect', and `fix'. 
% cite earlier work that does the search-for-keywords-to-identify-bugs; my earlier EMSE paper has that citation for instance. -- same issue as above, not sure how to cite

@@

% TODO fix this caption
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{figs/clusters.png}
\caption{\label{fig:clustering} Clustering results with nine different combinations of parameters ($\epsilon=[10^{-2}, 10^{-3}, 10^{-4}], Z=[5,7,9])$}
\end{figure}

@@

In this work, we were only interested in the first two groups and ignored the clusters for refactoring changes. Also, we prefer reporting clusters that manifest concrete patterns. An abstract pattern as opposed to a concrete pattern can account for many changes in the programs. For instance, the pattern `Changing a clone of a variable to a borrowing of it in a function argument' is deemed more concrete than `Adding a new statement to a function's body':

\begin{lstlisting}[language=Rust, style=colouredRust]
// First Pattern
-- foo1(arg.clone);
++ foo1(&arg);

// Second Pattern
fn foo2() {
    stmt1;
++  stmt2;
}

\end{lstlisting}

### 2022-09-23 sections/results.tex
added essence figure, revised methodology, added introduction for results, revised tables for results, we still need to add some numbers

@@

[explain why RQ1]

@@

@@

@@

\multirow{10}{*}{General}        
& \multirow{2}{*}{Attributes}    & G.attr.struct  & Modifying the attributes of structs                                        \\
                                 &                                & G.attr.struct.field  & Modifying the attributes of struct fields                                  \\\cline{2-3}
                                 & \multirow{2}{*}{Struct}        & G.struct.field  & Adding/Removing a struct field                                      \\
                                 &                                & G.struct.field.pub  & Modifying the access modifiers of struct fields     
                                  \\\cline{2-3}
                                 & Option                         & G.option  & Changing field type T to T$\langle$Option$\rangle$                     \\\cline{2-3}
                                 & \multirow{4}{*}{Types}         & G.type.field  & Changing a struct field type                                        \\
                                 &                                & G.type.generic  & Changing a generic type                                             \\
                                 &                                & G.type.enum.variant  & Change in enum variant value type                                   \\
                                 &                                & G.type.tuple & Changing the type inside a tuple                                         \\\cline{2-3}
                                 & \multirow{1}{*}{Traits}        & G.trait.bound & TraitBound Change                                               \\\cline{1-3}
\multirow{9}{*}{Borrow Checker} &
\multirow{2}{*}{Clone}                                         & BC.clone.ref & Dropping clone and adding borrowing                                                   \\
                                 &                                & BC.clone.drop & Dropping clone   
                                    \\\cline{2-3}
    & \multirow{3}{*}{Ref and Deref} & BC.ref.add  & Adding Borrowing                                                           \\
                                 &                                & BC.ref.drop  & Removing Borrowing                                                         \\
                                 &                                & BC.deref.to\_string  & Modification in dereferencing a string prior to calling to\_string()                                                            \\\cline{2-3}
                                 & \multirow{2}{*}{Mut}           & BC.mut.add  & Adding mutability                                                          \\
                                 &                                & BC.mut.drop  & Dropping mutability                                                        \\\cline{2-3}
                                 & Vector                         & BC.vec.slice  & Changing a Vec reference to Slice                                           \\\cline{2-3}
                                 & \multirow{1}{*}{Lifetime}      
                                                                 & BC.lifetime.static  & Dropping static lifetime

\end{tabular}
\caption{\label{fig:category}Bug Fix Pattern Categorization}
\end{table}

>>>>

In this section, we first evaluate the effectiveness of our data modelling approach (RQ1). That is, how successfull is our data modelling approach in capturing the most important aspects of a program change. This evaluation is required, as the credibility and generality of our clusters are heavily based on this step in our pipeline. Then we move to introducing the obtained clusters in two sections: Common patterns (RQ2), and BC-Related patterns (RQ3).

@@

@@

\multirow{10}{*}{General}        
& \multirow{2}{*}{Attributes} & G.attr.struct & -                                  \\
& & G.attr.struct.field & -                     \\\cline{2-3}
& \multirow{2}{*}{Struct} & G.struct.field  & -\\
& & G.struct.field.pub & -    
\\\cline{2-3}
& Option & G.option & -                    \\\cline{2-3}
& \multirow{4}{*}{Types} & G.type.field  & - \\
& & G.type.generic & -\\ & & G.type.enum.variant  & - \\
& & G.type.tuple & -                                         \\\cline{2-3}
& \multirow{1}{*}{Traits} & G.trait.bound & - \\ 
\hline
\end{tabular}
\caption{\label{table:general}Bug Fix Pattern Categorization (General Patterns)}
\end{table}

@@

\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\multirow{9}{*}{Borrow Checker} &
\multirow{2}{*}{Clone}                                         & BC.clone.ref & - \\
& & BC.clone.drop & -   
\\\cline{2-3}
& \multirow{3}{*}{Ref and Deref} & BC.ref.add & - \\
& & BC.ref.drop & - \\
& & BC.deref.to\_string  & -                                            \\\cline{2-3}
& \multirow{2}{*}{Mut} & BC.mut.add  & - \\
& & BC.mut.drop  & -                                \\\cline{2-3}
& Vector & BC.vec.slice  & -                                    \\\cline{2-3}
& \multirow{1}{*}{Lifetime}      
& BC.lifetime.static  & -
\\
\hline
\end{tabular}
\caption{\label{table:bc}Bug Fix Pattern Categorization (BC-Related Patterns)}
\end{table}

@@
